A Type System for String Sanitation\\Implemented Inside a Python

### 2014-04-13 sanitation-plas14/sanitation-plas14.tex

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
In web applications and other settings, incorrect input sanitation often causes
security vulnerabilities. For this reason, frameworks often provide methods
for sanitizing user input. When these methods are insufficient or unavailable,
developers often create  custom input sanitation algorithms. In both cases,
input sanitation techniques are ultimately implemented using the language's
regex library.

Formally, input sanitation is the problem of ensuring that an arbitrary string
is converted into a safe form before potentially unsafe use. For example, consider
SQL injection attacks. To prevent such attacks, we might ensure that any string
used as input to a query does not contain unescaped SQL command sequences.

This appendix presents a type system, $\lcs$, for ensuring that input 
sanitation algorithms are implemented correctly with respect to use site 
specifications.

Unlike frameworks provided by general purpose languages such as Haskell and
Ruby, our type system provides a \emph{static} guarantee that input is always 
properly sanitized before use. We achieve this by defining a typing relation
which captures idiomatic sanitation algorithms. Type safety relies upon 
several closure and decidability results about regular languages.

XDuce typechecks XML. Like our work, XDuce relies upon some properties of
regular expressions to establish soundness and completeness results. Unlike
XDuce, our work is motivated by input sanitation and therefore considers arbitrary
strings (as opposed to tree-structured XML documents). Furthermore, our static
treatment of the ubiquitous regex filter (replace) function for regular expression is novel.

Finally, some research languages achieve similar safety guarantees via a specialized type
system. The existence of such domain-specific type systems is suggestive, and 
we believe the simplicity of $\lcs$ demonstrates the effectiveness of
extensibility. Instead of introducing entirely new languages for each domain,
language extension developers may invest incrementally in obtaining static
guarnatees which address common mistakes.

>>>>

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
In web applications and other settings, incorrect input sanitation often causes
security vulnerabilities. In fact, ...OWASP says it's important... .
Mention CVE stats.
For this reason, modern web frameworks and libraries use various techniques to
ensure proper sanitation of arbitrary user input. On the rare occasions when
these methods are unavailable or insufficient, developers (hopefully) create 
ad hoc sanitation algorithms. Im most cases, sanitation algorithms -- ad hoc 
or otherwise -- are ultimately implemented using the language's regular
expression capabilities. Therefore, a system capable of statically 
checking properties about operations performed using regular expressions will
be expressive enough to capture real-world implementations of sanitation 
algorithms.

Input sanitation is the problem of ensuring that an arbitrary string
is coerced into a safe form before potentially unsafe use. For example, 
preventing SQL injection attacks requires ensuring that any string
coming from user input to a query does not contain unescaped SQL. 
The point at which arbitrary user input is concatenated into a SQL query is called a use site.
Although we believe are general approach extends to a wider class of problems
(e.g. sanitation algorithms for preventing XSS attacks might be definable using
regular tree languages), these generalizations are beyond the scope of the 
present paper.

This paper presents a language extension, definable in the Ace programming language,
for ensuring that input sanitation algorithms are implemented correctly with
respect to use site specifications. If use site specifications are sufficient, 
then type checking ensures the absence of vulnerabilities and other bugs which
arise from improper sanitation. Our extension focuses on sanitation algorithms 
which aim to prevent command-injection style attacks (e.g. SQL injection or RFI).

\subsection{Related Work and Alternative Approaches}

The input sanitation problem is well-understood. There exist a large number
of techniques and technologies, proposed by both practitioners and researchers, for
preventing injection-style attacks. In this section, we defend the novelty and
significance of our approach with respect to the state of the art in practice 
and in research.

Unlike frameworks provided by languages such as Haskell and
Ruby, our type system provides a \emph{static} guarantee that input is always 
properly sanitized before use. We achieve this by defining a typing relation
which captures idiomatic sanitation algorithms. Type safety in our system relies upon 
several closure and decidability results about regular languages.

Libraries and frameworks available in functional programming language communities often
make claims about security and sometimes even sophistically mention sophisticated
type systems as evidence of freedom from injection-style attacked.
However, even the specification that input is always sanitized properly before use is not
actually captured by the type system or anywhere else; in fact, the full specification of these
algorithms is rarely characterized by anything more specific than the type
String $\rightarrow$ String, which is a class of safe input sanitation algorithms
only in the most degenerate case.

A number of research languages provide actually static guarantees that a program
is free of input sanitation vulnerabilities. Most rely on some form of information
flow. TODO-nrf  give citations and examples. Our extension to Ace differs from
these systems in the ways following:
\begin{itemize}
  \item Our system is a light-weight solution to a single class of sanitation
  vulnerabilities (e.g. we do not address Cross-Site Scripting). We present our
  system not as a comprehensive solution to the web security problem, but rather as
  evidence that composable, light-weight and simple analyses can address security
  problems.
  \item Our system is defined as a library in terms of an extensible type system,
  as opposed to a stand-alone language. This is important for two reasons.
  First, our system is one of the first large examples written in Ace, and serves
  as an extended case study. Second, although our approach requires
  developers to adopt a new language, it does not require developers to adopt
  a language specifically for web development.
  \item Ace is implemented in Python and shares its grammar. Since Python is a 
  popular programming language among web developers, the barrier between our research
  and adopted technologies is far lower than is e.g. Ur/Web's.
  \item In general, our is a composable, light-weight and
  natural solution to a single problem -- rather than a comprehensive 
  solution to the web security problem. Our goal is to demonstrate that extensible
  type systems can capture static properties of common idioms, and thereby ensure
  safety without introducing much additional complexity.
\end{itemize}

Finally, incorporating regular expressions into the type system is not novel.
The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
We differ from this and related work in at least two ways.
First, our system  is defined within an extensible type system;
this first difference introduced an interesting class of design problems (see \S ???).
TODO-nrf this is the second time this is used as a warrant. Factor out this argument
and place it at the top???
Second, our focus on security required novel design decisions; for instance, the
filter elimination form is unique to Ace.

In conclusion, our system is novel in at least two ways:
\begin{itemize}
\item The safety guarantees provided by libraries and frameworks in popular languages
are not as (statically) justified as is often belived (or even claimed).
\item Our extension is the first major demonstration of how an extensible type
system may be used to provide light-weight, composable security analyses based upon
idiomatic code.
\end{itemize}

\subsection{Outline}
